---
title: "The NLSY79 Wages Data of High School Dropouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The NLSY79 Wages Data of High School Dropout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r library}
library(yowie)
library(dplyr)
library(ggplot2)
library(ggmosaic)

```

This vignette provides a brief example of the usage of `wages_hs_dropuot` data. 

## Demographic Profile of High School Dropouts

Here, we would like to visualize the demographic profile of high school dropout cohort in the NLSY 1979 by gender and race. 


```{r}
ggplot(wages_hs_dropout) +
  geom_mosaic(aes(x = product(race),
                  fill = gender)) 
```

## Comparing The Wages of Dropouts Group and Other Groups

In this section, we would like to compare the summary statistics of high school dropouts , those who completed high school without being dropped out, and those who never attended high school.

**High school dropouts cohort**

```{r}
summary(wages_hs_dropout$mean_hourly_wage)
```

**Completed high school without being dropped out cohort**

```{r}
hs_never_dropout <- wages_hs2020 %>%
  mutate(dob = 1979 - age_1979,
         age_hgc = yr_hgc - dob) %>%
  filter(hgc == "12TH GRADE" & age_hgc < 19)
summary(hs_never_dropout$mean_hourly_wage)
```

**Never attended high school cohort**
```{r}
not_attend_hs <- wages_hs2020 %>%
  filter(hgc_i < 9)
summary(not_attend_hs$mean_hourly_wage)
```



