---
title: "Demographic Data of The NLSY79 Cohort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demographic Data of The NLSY79 Cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(yowie)
library(dplyr)
library(ggplot2)
```

The aim of this vignette is to provide an example of how the demographic data that is provided in the package could be analyzed. 

## Profile of The NLSY79's Subjects

In this example, we would like to plot the distribution of the cohort by their race and gender. 

```{r}
ggplot(demographic_nlsy79) +
  geom_bar(aes(x = race,
               fill = gender),
           position = "dodge") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Join the Demographic Data and Wages Information

In this example, we want to observe whether there is a difference of the median of mean hourly wage of the high school cohort from different gender. We can do this by firstly calculate the median of mean hourly wage in `wages_hs2020` table and join it with `demographic_nlsy79`.

```{r}
med_wages <- wages_hs2020 %>%
  group_by(id) %>%
  summarise(median_wages = median(mean_hourly_wage))

med_wages_hs <- left_join(med_wages, 
                          demographic_nlsy79,
                          by = "id")

ggplot(med_wages_hs) +
  geom_density(aes(x = median_wages, 
                   colour = gender,
                   fill = gender),
               alpha = 0.3)
```

