---
title: "Wages Data from The NLSY79"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wages Data from The NLSY79}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r libraries}
library(yowie)
library(dplyr)
library(brolgar)
library(ggplot2)
library(patchwork)
library(tidyr)
```


The  aim of this vignette is to provide examples of the usage of wages data in longitudinal data analysis. In this case, we use `brolgar`, an R package, to analyze and visualize longitudinal data (Tierney, Cook, and Prvan, 2020).

## Explore the Data Feature

`brolgar` can only be used if the data frame is in a `tsibble` form. Hence, we should transform the data into this object first. The `key` would be the `id` of the subject, and the `index` is the `year` variable.  

```{r tsibble}
wages_hs2020 <- as_tsibble(x = wages_hs2020,
                    key = id,
                    index = year,
                    regular = FALSE)
```


### Visualize the number of observation per ID 

Using `brolgar`'s `features` function, we can extract the number of observations of each ID and plot it using `ggplot2`.  

```{r}
wages_hs2020 %>% 
  features(mean_hourly_wage, n_obs) %>%
  ggplot(aes(x = n_obs)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(5, 28, 3)) +
  theme_bw() +
  xlab("Number of Observations") +
  ylab("Count of ID")
```

We can learn that the subjects have different number of observations, however most of them completed 27 rounds of surveys. 

### When is the first and the last time the subject attend the survey?

We can also explore the distribution year of the IDs' first time and last time participated in the survey. 

```{r first-last-plot}

 first_time <- wages_hs2020 %>%
  features(year, list(min = min)) %>%
  ggplot(aes(x = min)) +
  geom_histogram(binwidth = 1) +
  xlab("First time in study") +
  scale_x_continuous(breaks = seq(1979, 2108, 5)) +
  theme_bw()

last_time <- wages_hs2020 %>%
  features(year, list(max = max))%>%
  ggplot(aes(x = max)) +
  geom_histogram(binwidth = 1) +
  xlab("Last time in study") +
  scale_x_continuous(breaks = seq(1979, 2108, 5)) +
  theme_bw()

first_time + last_time
```

Most of the subjects participated in the first round of the survey, which was in 1979. They also keep participating in the study until the last round the survey held, which was in 2018. 

### Summarize the Data Feature

We can also summarize the mean hourly wage using the `features` function. Here, we want to display the minimum, median, and maximum value of mean hourly wage of each ID. 

```{r three-feat}
wages_three_feat <- wages_hs2020 %>%
  features(mean_hourly_wage, 
           feat_three_num
           )

wages_feat_long <- wages_three_feat %>%
  pivot_longer(c(min, med, max), names_to = "feature", values_to = "value")
  
ggplot(wages_feat_long) +
  geom_density(aes(x = value, colour = feature, fill = feature), alpha = 0.3) +
  theme_bw() 
```

The distribution of min, median, and a max of the mean hourly wage is right-skewed, which means that some subjects earned too high hourly wages. Further, some IDs have minimum wages that are higher than others' maximum wages.

## Sample Some Subjects and Visualize It

Creating a line plot of the wages of all IDs might be too messy since there are 5931 IDs. Thus, we can take some sample of ID's randomly using `brolgar`'s `facet_sample` function. 

```{r sample}
set.seed(202102)

ggplot(wages_hs2020, 
       aes(x = year,
                y = mean_hourly_wage,
                group = id)) +
  geom_line(alpha = 0.7) +
  facet_sample() +
  ylab("mean hourly wage") +
  theme_bw()
```

From the plot, we can learn that the mean hourly wages fluctuated over the time, however, generally it increased since 1979. 






